Visualization and analysis of probability distributions of large temporal data

### 2021-11-04 thesis.Rmd
remove last 2 commits

---
title: 'Visualization and analysis of probability distributions of large temporal data'
author: 'Sayani Gupta'
degreetype: 'Doctor of Philosophy'
site: bookdown::bookdown_site
degrees: 'M.Stat, B.Sc (Stat Hons)'
bibliography: [bib/gravitas.bib, bib/hakear.bib, bib/gracsr.bib, bib/thesisrefs.bib]
link-citations: yes
---

@@

This work is driven by the need for a conceptual framework to tackle temporal analyses in a data-centric workflow. Most research on temporal data focuses on modeling. Corresponding software requires very stringently formatted data, but does little in providing guidelines or tools for wrangling raw data into the required format. This has led to ad-hoc, and once-off solutions, which this research repairs.

There are three original contributions for the temporal data analysis in this research. They are grounded in the spirit of exploratory data analysis for time-indexed data. The first contribution (Chapter \@ref(ch:gravitas)) is a new technique for visualizing data using a calendar layout. It is most useful when the data relates to daily human activity, and patterns can be explored in relation to people's schedules. The second contribution (Chapter \@ref(ch:hakear)) is a new data abstraction which streamlines transformation, visualization, and modeling in temporal contexts. This "tsibble" object is a data infrastructure forming the foundation of temporal data pipelines. The third contribution (Chapter \@ref(ch:gracsr)) is to equip analysts with exploratory and explanatory tools for discovering missing patterns in time, and thus facilitating decisions on appropriate imputation methods for further downstream analysis.

@@

@@

This thesis would not be possible without many people.

First and foremost, I would like to express my gratitude to my supervisors, Dianne Cook and Rob J Hyndman. They are outstanding supervisors, providing excellent insights on statistical computing and graphics during our weekly supervisory meetings. I'm deeply impressed by their dynamics, enthusiasm, and visions about open source software and data science, all of which have formed the support and encouragement for me to continue with this non-traditional research path.

I thank Heike Hofmann for hosting me at Iowa State University during the summer of 2018. I enjoyed many discussions with Heike, where I received brilliant questions and insightful advice on my research. I thank Stuart Lee and Mitchell O'Hara-Wild for countless conversations on software development, that helped to shape this thesis. I thank David Frazier for being the best pingpong partner, truly boosting the creative thinking process.

I thank Emi Tanaka for sending me PRs for proofreading my thesis on Github, unlocking the power of an open-sourced research compendium. I thank Wenjing Wang and Carson Sievert for being so helpful in the early days of my PhD. I thank the R community, the tidyverse team, and all NUMBATs for ongoing inspiration.

Last but not least, I'd like to thank my mother for her unceasing love.

@@

# Data representation, visual and analytical techniques for demystifying temporal missing data {#ch:mists}

@@

## Software development
### sugrrants
### tsibble
### mists
## Future work
### Process for generating missing data in time
### Visual methods for temporal data of nesting and crossing interactions
## Final words

>>>>

---
title: "Visualization and analysis of probability distributions of large temporal data"
author: "Sayani Gupta"
degreetype: Doctor of Philosophy
site: bookdown::bookdown_site
degrees: M.Stat, B.Sc (Stat Hons)
bibliography:
- bib/gravitas.bib
- bib/hakear.bib
- bib/gracsr.bib
- bib/thesisrefs.bib
link-citations: yes
---

@@

The research was motivated by the desire to understand the Australian smart meters data, which was collected half-hourly for two years at the household level. With temporal data available at ever finer scales, exploring periodicity can become overwhelming with so many possible temporal deconstructions to explore. Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming.

<!-- exploration may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process. It is important to be able to navigate through all the temporal deconstructions that accommodate for periodicities to have multiple perspectives of the observed data. This idea aligns with the notion of EDA (Tukey 1977) which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Slicing and dicing the data in all possible temporal scales as suggested by EDA can be a daunting task as it leads to a -->
<!-- myriad of possibilities. -->

<!-- Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming -->

<!-- exploring periodicity can become overwhelming with so many possible temporal deconstructions to explore. Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming. -->

<!-- <WHY> <WHAT> <HOW> -->

This work investigates how time may be dissected, resulting in alternative data segmentation and, as a result, different visualizations that can aid in the identification of underlying patterns. The first contribution (Chapter \@ref(ch:gravitas)) describes classes of time deconstructions using linear and cyclic time granularities. It provides tools to compute possible cyclic granularities from an ordered (usually temporal) index and also a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities by defining "harmony". "harmony" denotes pairs of granularities that could be analyzed together and reduces the search from all possible options. This approach is still overwhelming for human consumption because there would still be a huge number of harmonies. Hence, the second contribution (Chapter \@ref(ch:hakear)) refines the search of informative granularities by identifying those for which the differences between the displayed distributions are greatest and rating them in order of importance of capturing maximum variation. The third contribution (Chapter \@ref(ch:gracsr)) builds upon the first two to provide methods for exploring heterogeneities in repetitive behavior for many households and over multiple granularities. It accomplishes this by providing a way to cluster time series based on probability distributions across informative cyclic granularities. Although we were motivated by the smart meter example, the problem and the solutions we propose are practically relevant to any temporal data observed more than once per year.

<!-- there is a need to resort to clustering, which has been addressed in  It approaches the problem by computing pairwise distances between time series based on probability distributions over multiple cyclic granularities at once. -->
\newpage

@@

I have not renumbered sections of submitted or published papers in order to generate a consistent presentation within the thesis.

\textbf{Student name:} Sayani Gupta

\textbf{Student signature:}

```{r signature, out.width = "20%", fig.align = "left", eval = file.exists("img/signature.png")}
knitr::include_graphics("img/signature.png")
```

\textbf{Date:} `r Sys.Date()`

@@

## supervisors
Rob and Di are phenomenal leaders in their respective fields, who lead by example. I am inspired by their creativity, wisdom, discipline, and dedication to really contribute to the society through their research. Thank you for constantly pushing me to improve as a programmer and researcher, and for regularly sharing best practices for conducting research. Looking back, I am grateful for how my thoughts and work in statistical computing, graphics, and data analytics in general have evolved over the years. I still have a lot to learn, but I am a much more self-reliant and independent researcher than I was when I started. Di has been instrumental in exposing me to the potential benefits of effective data visualization. As a female researcher, I am encouraged by her willingness to pursue unconventional avenues and have frequently noticed her make a conscious decision to question existing  stereotypes and biases. Thank you Di for being a fantastic female role model.

## department and PhD colleagues

## R community

## friends

PhD in Covid has been harder and I would not have been able to make it through this if not for the emotional support of my friends and family. Thank you Puwasala, Sium for being an epitome of kindness and acting as a pillar of support. Thank you Tushar, Samarpita and Nairita for always motivating me and believing in me when I didn't believe in myself. Thank you Ian for always putting things in perspective when I lost sight. Thank you each one of you for being available, listening to me, and not giving up on me at my lowest. Thanks to my housemates Anjali, Surbhi and Dulaji for the fun company and conversations that kept me sane in the covid lockdowns.

## family
Finally, a big thanks to my family for always being supportive of my choices. Thanks to my mum (Nupur Gupta) and dad (Arun Prasad Gupta), from whom I learnt that no matter where you start from, if you persevere and are sincere in your efforts, you can sail through any difficult situation. They are my constant cheerleaders for all little and big endeavours. Thanks to my brother (Avijit) for always having my back and inspiring me to dream bigger. Thank you, Juhi (sister-in-law) for instilling in me the value of organisation in all aspects of life and the importance of prioritizing my physical health from time to time. Thank you all for your part in supporting me in this journey. I love you.

@@

# Clustering time series based on probability distributions across temporal granularities {#ch:gracsr}

## Introduction
## Clustering methodology {#sec:methodology}
## Validation {#sec:validation}
### Data generating processes
### Data designs  
#### Individual granularities
#### Interaction of granularities
### Visual exploration of findings
## Case study {#sec:cases}
### Clustering
#### JS-based distances
#### wpd-based distances
## Conclusion and future work {#sec:conclusion}
## Acknowledgments {-}
## Supplementary Materials

@@

## Software development
### gravitas
### hakear 
### gracsr
## Future work
### Putting all functionalities on CRAN
### Scaling up the clustering method to incorporate large uncertainty and improved computational efficiency that comes with large data
### Comparing our clustering method with other benchmark methods
### Check generalizability of our methods
## Final words
